---
title: "Problem Statement EDA Analysis"
author: "Riyas KP"
date: "6/24/2019"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

### __Introduction to Problem Statement__ 

You work for a __consumer finance company__ which specializes in lending various types of loans to urban customers. When the company receives a loan application, the company has to make a decision for loan approval based on the applicant’s profile. __Two types__ of risks are associated with the bank’s decision:

* If the applicant is likely __to repay the loan__, then not approving the loan results in a __loss of business to the company__

* If the applicant is __not likely to repay the loan__, i.e. he/she is likely to default, then approving the loan may lead to a __financial loss__ for the company

The data given below contains the information about past loan applicants and whether they ‘defaulted’ or not. The aim is to identify patterns which indicate if a person is likely to default, which may be used for taking actions such as denying the loan, reducing the amount of loan, lending (to risky applicants) at a higher interest rate, etc.

In this case study, you will use EDA to understand how __consumer attributes__ and __loan attributes__ influence the tendency of default.

When a person applies for a loan, there are two types of decisions that could be taken by the company:

1. __Loan accepted__: If the company approves the loan, there are 3 possible scenarios described below:
* __Fully paid__: Applicant has fully paid the loan (the principal and the interest rate)

* __Current__: Applicant is in the process of paying the instalments, i.e. the tenure of the loan is not yet completed. These candidates are not labelled as 'defaulted'.

* __Charged-off__: Applicant has not paid the instalments in due time for a long period of time, i.e. he/she has __defaulted__ on the loan

2. __Loan rejected__: The company had rejected the loan (because the candidate does not meet their requirements etc.). Since the loan was rejected, there is no transactional history of those applicants with the company and so this data is not available with the company (and thus in this dataset)

### __Business Objectives__

This company is the largest online loan marketplace, facilitating personal loans, business loans, and financing of medical procedures. Borrowers can easily access lower interest rate loans through a fast-online interface.

Like most other lending companies, lending loans to ‘risky’ applicants is the largest source of financial loss (called credit loss). The credit loss is the amount of money lost by the lender when the borrower refuses to pay or runs away with the money owed. In other words, borrowers who __default__ cause the largest amount of loss to the lenders. In this case, the customers labelled as 'charged-off' are the 'defaulters'.

If one is able to identify these risky loan applicants, then such loans can be reduced thereby cutting down the amount of credit loss. Identification of such applicants using EDA is the aim of this case study.

In other words, the company wants to understand the __driving factors (or driver variables)__ behind loan default, i.e. the variables which are strong indicators of default. The company can utilize this knowledge for its portfolio and risk assessment.

## __Analysis__

### __Overview of the Dataset__

__Browsing through the Data:__

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code. 
# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(ggplot2)
theme_set(theme_minimal(12))
```

```{r echo=FALSE, loding_the_data}
#importing the loan file and assigning to a varible called loan
loan <- read.csv('loan.csv')
```

This data set contains 39,717 loan information with 111 variables on each loan.

```{r echo=FALSE, dimention_of_the_data}
#checking how many varibles and the total number of observation are recorded for each variable
dim(loan)
```

Here are the names of the 111 variables:

```{r echo= FALSE, names_of_variables}
#to check the names of varibles
names(loan)
```

Lets have a look at the structure of the dataset:

We can see that there are both quatitative and categorical variables, quantitaive variales are numerical numbers and Factor is a categorical varible, right next to the Factor you can identify how many categorical levels are there in the specified vatiable.

```{r echo= FALSE, structure_of_the_data}
#to check the type of observation stored in a varible
str(loan)
```

Considering the below table of application_type varibale, We can identify that there are 39717 inidvidual applications.

```{r echo= FALSE, type_of_application}
#this code will group the number of different observation for the variable application_type
table(loan$application_type)

```

### __Univariate Analysis__

Lets have look into the amount of the loan applied for by the borrower.

```{r echo= FALSE, loan_amnt_histogram}
#creating a plot of loan amount using the ggplot
ggplot(aes(x= loan_amnt), data= loan) +
  geom_histogram(binwidth = 750, color = 'black', fill = '#099DD9') +
  xlab('The loan amount applied by the borrower') +
  ylab('Proportion of users')
```

The number of payments on the loan in months.

```{r echo =FALSE,warning=FALSE, months_plot}
#creating a plot of terms using ggplot
ggplot(aes(x= term), data= loan) +
  geom_histogram(stat="count", color = 'black', fill = '#D8BFD8') +
  xlab('Number of payments(Values are In months)') +
  ylab('Proportion of users')
```

From the above observations we can clearly understand that the most of the borrowed amount is near to $10,000. And most chosen number of payment is 36 months which is three years.

Lets have a look into the status of home ownership:

```{r echo=  FALSE,warning=FALSE, home_ownership_status}
#creating a plot using of ownership status using ggplot
ggplot(aes(x= home_ownership), data= loan) +
  geom_histogram(stat="count", color = 'black', fill = '#FFDAB9') +
  xlab('Home ownership status') +
  ylab('Proportion of users')
```

From the above plot we understood that most of the people are renting the their property and ther is also much similar number of people who are on a mortgage. 


Below plot will show when does a person get a loan in terms of the length of their working status.

```{r echo= FALSE, warning=FALSE, employee_length_plot}
#creating a plot of employee length using ggplot
ggplot(aes(x= emp_length), data = loan) +
  geom_histogram(stat = "count", color = 'black', fill = '#00BFFF') +
  xlab('Employment length in years') +
  ylab('Proportion of users')
```

Tabel below is showing the grouped summary of the employment length.

```{r echo= FALSE, employee_length}
#grouping the number of different observation for the employment length into a table
table(loan$emp_length)
```

In the above observations we can make out that there are 1075 loans with no recorded employment status, these number shows us that either 1075 number of people are unemployed or we have missed the information of theire status of employment. This plot also shows that a good number of people with more than ten years of employment are taking loans, which is almost a 23%. And 12% of people who took loans are less than a year employed. During First year and fifth year we see almost similar number of loans. Second year we have almost similar number of loans compare to the less than a year employed people. from the third year onwards we see a decreasing trend till nineth year.


Lets have a look at the annual income of the borrowers:

```{r echo = FALSE, warning=FALSE, income_plot}
#creating a plot of annual income using ggplot
#i have used a log based scale to get an efficient x axis measure
ggplot(aes(x= annual_inc), data = loan) +
  geom_histogram(binwidth = .1, color = 'black', fill = '#DEB887') +
  scale_x_log10() +
  xlab('Self reported annual income in log10') +
  ylab('Proportion of users')
  
```

Below is the summary of the annual income provided by the borrower.

```{r echo = FALSE}
#to get a summary of annual income variable
summary(loan$annual_inc)
```

The income of most borrowers range from $40,000 to $82,000, the highest recorded income is $6,000,000.


The below plot will give us a picture of the current staus of the loan:

```{r echo=  FALSE, warning=FALSE, loan_status_plot}
#creating a plot of loan status using ggplot
ggplot(aes(x= loan_status), data = loan) +
  geom_histogram(stat = "count", color = 'black', fill = '#FFB6C1') +
  xlab('Current status of the loan') +
  ylab('Proportion of users')

```

Grouped loan status table.

```{r echo= FALSE, loan_status_table}
#grouping the number of different observation for the current loan status into a table
table(loan$loan_status)
```

From the above observation we can make out that there are 5627 loans which had been defaulted. Which is a 14% of total number of loan given.

Lets see for what purpose people borrow money for:

```{r echo= FALSE, warning=FALSE, loan_purpose_plot}
#creating a plot for purpose of the loan using ggplot
ggplot(aes(x= purpose), data =loan) +
  geom_histogram( binwidth = 1, stat =  'count', color = 'black', fill ='#FFA07A') +
  xlab('Category provided by the borrower') +
  ylab('Proportion of users')
```

Below table give us the exact number of different loan purpose.

```{r echo= FALSE, purpose_table }
#grouping the observations purpoe of the loan into a table
summary(loan$purpose)
```

Majority of the people got the loan for debt consolidation. Credit card is also an important purpose of the loan. Another thing I found is about somewhat similar amount of people got loans for home improvement and major purchase. One interesting thing i found that few number of people are taking loan for renewable energy. This is a good sign that people are thinking about the climate change and they are acting. We need to have a deeper understanding of this and need to think about how can we encourage more people apply for taking loan for renewable energy. May be we need to consider some special interest rates for this in the future. And a good portion of the people didnt mentioned the specific purpose of the loan.


Lets have a look into the Debt to income ratio of borrowers:

```{r echo= FALSE, warning=FALSE, dti_plot}
#creating a plot of debt to income ratio using ggplot
ggplot(aes(x=dti), data = loan) +
  geom_histogram(binwidth = .5, color = 'black', fill = '#9370DB') +
  scale_x_continuous(breaks = seq(1,32,1)) +
  xlab('Debt to income ratio') +
  ylab('Proportion of users')

```

Inorder to find out how many people are having dti <1, we will use below quantile function. 


```{r echo =  FALSE}
#quantile produces sample quantiles corresponding to the given probabilities
quantile(loan$dti,0.99)
```

After limiting the ratio to <1, we can understand that only 29% has debt to income ratio less than 1. And if we call the function quantile(loan$dti, 0.99, na.rm=T), it tells us that almost 75% of the borrowers has their dti > 1. Which is dnagerously high. Evidence from studies of loans suggest that borrowers with a higher debt to income ratio are more likely to run into trouble making monthly payments.


Income Verification:

```{r echo=  FALSE,warning=FALSE,  verificatio_status_plot}
#creating a plot of verification status using ggplot
ggplot(aes(x= verification_status), data = loan) +
  geom_histogram(stat = "count", color = 'black', fill = '#FAFAD2') +
  xlab('Verification status of the loan') +
  ylab('Proportion of users')
```

```{r echo = FALSE, verificaito_status_table}
table(loan$verification_status)
```

From the above plot we can understand that half of the loan was allowed wihout any verification. Approving a loan without verifing documents and backgroudnd is also a cause to default.

### __Bi-Variate Anlysis__

Lets have a look into the annual income and monthly installment of the borrowers.

```{r echo= FALSE, annual_inc&installment}
#creating a scatter plot using ggplot and i have used log based scale for a better
#measurement of x axis
ggplot(aes(x = annual_inc, y = installment), data = loan) +
  geom_point() +
  scale_x_log10() +
  xlab('Annual income of the borrower') +
  ylab('Monthly payment owed by the borrower')
```

From the above plot we can identify a  positive moderate upward curve from the obervations. It shows us the streangth and direction of  the varibles. If we verify annual income of all the borrowers, we dont have to do anything new with the monthly payment calculation for each loan.


Loan amount and debt to income ratio plot:

```{r echo= FALSE, loan_amnt&dti}
#creating a scatter plot using ggplot
ggplot(aes(x = loan_amnt, y = dti), data = loan) +
  geom_point() +
  xlab('Amount of the loan applied for by the borrower') +
  ylab('Debt to income ratio')
```

The above plot is giving us a weak spread out picture of observations, it shows us how inappropiate and high debt to income ratio was when we comapare it to the loan amount had taken by the borrower.

```{r echo= FALSE, loan&income_scatterplot}
ggplot(aes(x = loan_amnt, y = annual_inc), data = loan) +
  geom_point() +
  scale_y_log10() +
  xlab('Amount of the loan applied for by the borrower') +
  ylab('Annual income of the borrower')
```

The above plot is reflecting that the points are spreaded in a slight moderate positive upward direction. which showing us that the people having higher annual income takes highest loan amount. The higher number of loans are from people who are having an annual income of $10,0000 and less, which indicates the wide spread dark concentrated points in the plot. Vertical bar shows us that most people takes a rounded figure like, $10k, $15k, $20k and so on.


### __Final Plots and Summary__

1. __Income Verification:__

```{r echo=  FALSE,warning=FALSE,  verificatio_status_plot_final}
#creating a plot of verification status using ggplot
ggplot(aes(x= verification_status), data = loan) +
  geom_histogram(stat = "count", color = 'black', fill = '#FAFAD2') +
  xlab('Verification status of the loan') +
  ylab('Proportion of users')
```

Verfication is a keyfactor, and income verification has to be done in a way that there is no forgery involved. If the annual incomes of all the borrowers are verified then only we will be able to get a proper debt to income ratio.  


2. __Debt to income ratio:__

```{r echo= FALSE, warning=FALSE, dti_plot_final}
#creating a plot of debt to income ratio using ggplot
ggplot(aes(x=dti), data = loan) +
  geom_histogram(binwidth = .5, color = 'black', fill = '#9370DB') +
  scale_x_continuous(breaks = seq(1,32,1)) +
  xlab('Debt to income ratio') +
  ylab('Proportion of users')

```

Debt to income ratio is the monthly debt payments divided by gross monthly income. This number is one way lenders measure borrowers ability to manage the payments they make every month to repay the money borrowed. 

We understood that only 26% has debt to income ratio less than 1. It shows us that almost 75% of the borrowers has their dti > 1. Which is dnagerously high. Evidence from studies of loans suggest that borrowers with a higher debt to income ratio are more likely to run into trouble in making monthly payments.Higher risk is usually associated with higher monthly payment. And borrowers must keep their debt to income ratio low. Paying a huge amount of money from monthly salary in every month for a period of 3 years is not possible, of course you have other things to do in daily life, you need money for that also.


3. __Loan amount and debt to income ratio plot:__

```{r echo= FALSE, loan_amnt&dti_final}
#creating a scatter plot using ggplot
ggplot(aes(x = loan_amnt, y = dti), data = loan) +
  geom_point() +
  xlab('Amount of the loan applied for by the borrower') +
  ylab('Debt to income ratio')
```

The above plot is giving us a weak spread out picture of observations, it is a reflection of how inappropiate and high debt to income ration was when we comapare it to the loan amount had taken by the borrower.

                                                                            End

__Reflection__

I have created this markdown dcoument in R language inside RStudio. I have used ggplot to create different types of plots and visualization, i found it simple and efficient using ggplot rather than the defualt R syntax.

Since i have been learnig R for sometime, this is my very first interaction with a live project.
Thank you for showing interest and sharing the information.
